Dependent Object Types

### 2012-09-22 doc/fool/article.tex
coq cite and minor rev
more elaborate details on why the patch failed. copied from report, mostly.

@@

We believe that the DOT calculus is type-safe, and are developing a
proof of type safety using step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In this section, we
briefly summarize the main theorem of type safety, and the strategy
based on step-indexed logical relations that we are using to prove
it. All our development, including models in Coq and PLT Redex, is
available from \url{http://dot.namin.net}.

@@

We tried! However, the resulting calculi were not elegant, and
furthermore, we still found issues with preservation. Below, we give a
brief summary of one failed attempt to patch the calculus for
preservation to hold.

Because many of the counterexamples to preservation are related to
narrowing, we tried to make widening an explicit operation and change
rules with implicit relaxations (\textsc{msel} and \textsc{new}) to be
strict. From a typing perspective, the change was straightforward, but
reduction became more complicated and dependent on typing because the
type information in widenings needed to be propagated correctly.

We added path equality provisions in the subtyping rules, in the same
spirit as the Tribe calculus~\cite{tribe}.

Unfortunately, these two patches interacted badly, and we were left
with a disturbing counterexample to type safety.

\begin{align*}
&\mlnew a {\Top \refine z {\Ldecl C \Bot {\Top \refine z {\Ldecl D \Bot {z.X}, \Ldecl X \Bot \Top}}}}{
\mlnew b {{a.C} \refine z {\Ldecl X \Bot \Bot}}}{
\mlnew c {a.C}}{
\mlnew d {{\cast {a.C} b}.D}}{
\app {(\abs x \Bot {x.\mathit{foo}})} d
}
\end{align*}

Notice that $d$ has type $\Bot$ if the cast on $b$ is ignored. This
example didn't typecheck initially because the path-equality
provisions only applied when objects are in the store, so the
application was not well-typed. But if we started preservation in a store
which had $a$, $b$, $c$ and $d$ then the application type-checked,
because, through one of the path-equality provision, we could find that
the type of $d$ was a subtype of $\Bot$. Now, of course, when we got to
$d.\mathit{foo}$, reduction failed.

@@

Proving the DOT calculus type-safe has been an interesting
adventure. We have shown that DOT does not satisfy preservation (also
known as subject-reduction). However, the standard theorems of
preservation and progress are just one way to prove type safety. We
have sketched a plausible proof of type safety using the powerful
method of logical relations.

>>>>

\subsection{Expansion}\label{ex-exp}

@@

We believe that the DOT calculus is type-safe, and are developing a
proof of type safety using step-indexed logical
relations~\cite{ahmed04,ahmed06,step_indexed_obj}. In this section, we
briefly summarize the main theorem of type safety, and the strategy
based on step-indexed logical relations that we are using to prove
it. All our development, including models in Coq~\cite{coq} and PLT Redex~\cite{plt_redex}, is
available from \url{http://dot.namin.net}.

@@

We tried! However, the resulting calculi were not elegant, and
furthermore, we still found issues with preservation.

Here is a summary of one attempt, which is further detailed
below. Because many of the counterexamples to preservation are re-
lated to narrowing, we tried to make widening an explicit operation
and change rules with implicit relaxations (MSEL and NEW) to be
strict. From a typing perspective, the change was straightforward, but
reduction became more complicated and dependent on typing because the
type information in widenings needed to be propagated correctly.  We
added path equality provisions in the subtyping rules, in the same
spirit as the Tribe calculus~\cite{tribe}.  Unfortunately, these two
patches interacted badly, and we were left with a disturbing
counterexample to type safety.

In any case, this attempt resulted in a patched calculus that was not
as elegant as the original one, in addition to being unsound. We're
open to ideas!

In the DOT calculus as presented, the \textsc{app} and \textsc{new} typing
rules have implicit relaxations. For instance, in \textsc{app}, the
argument type may be a subtype of the declared parameter type. In
order to deal with all the preservation counterexamples due to narrowing, we
tried making widening an explicit operation and changing those rules to be
strict by replacing those relaxed subtyping judgments with equality
judgments. Two types $S$ and $T$ are judged to be equal if $S <: T$
and $T <: S$.

Syntactically, we add a widening term: $\wid t T$, and extend values
with a case for widening: $\wid v T$. The typing rule for widening,
\textsc{wid}, is then the only one admitting a subtyping relaxation:
$\Gamma \ts (\wid t T) \typ T$ if $\Gamma \ts t \typ T'$ and $\Gamma \ts T'
\sub T$.

The reduction rules become more complicated because the type
information in the widening needed to be propagated correctly. We will
motivate this informally with examples.

\begin{align*}
&\mlnew v {\Top \refine z {\Ldecl {L_a} \Bot \Top, \ldecl l {\Top \refine z {\Ldecl {L_a} \Bot \Top}}}\\
&\gap\gap\ldefs{l = \wid v {\Top \refine z {\Ldecl {L_a} \Bot \Top}}}}{
\app {(\abs x \Top x)} {(\wid v {\Top \refine z {\ldecl l \Top}}).l}
}
\end{align*}

The term $(\wid v {\Top \refine z {\ldecl l \Top}}).l$ first widens
$v$ so that the label has type $\Top$ instead of $\Top \refine z
{\Ldecl {L_a} \Bot \Top}$.

How should reduction proceed? We cannot just strip the widening and
then reduce, because then the strict function application would not
accept the reduced term. In short, we need to do some type
manipulations during reduction, by using the membership and expansion
judgments. This is a bit unfortunate, because it means that reduction
now needs to know about typing.

Next, we look at path equality provisions. These are even more
essential now in the presence of explicit widening. Consider this
example:

\begin{align*}
&\mlnew b {\Top \refine z {\Ldecl X \Top \Top, \ldecl l {z.X}} \ldefs{l = \wid b {b.X}}}{
\mlnew a {\Top \refine z {\ldecl i {\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}}\\
&\gap\gap\ldefs{i = \wid b {\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}}}}{
\wid {a.i.l} \Top
}
\end{align*}

$a.i.l$ reduces to $\wid b {\Top \refine z {\Ldecl X \Bot \Top, \ldecl
    l {z.X}}}$. Now, how can we continue? $b.l$ reduces to $b : b.X$
which has bounds $\Top..\Top$, but $(\wid b {\Top \refine z {\Ldecl X
    \Bot Top, \ldecl l {z.X}}}).l$ has bounds $\Bot..\Top$, so without
some provision for path equality, we cannot widen $b.l$ to $(\wid b
{\Top \refine z {\Ldecl X \Bot \Top, \ldecl l {z.X}}}).l$.

We add the path equality provisions to the subtyping rules.

Let's first ignore the extension of the calculus requiring explicit
widenings introduced in~\ref{widening}. Then, we need to add one
intuitive rule to the subtyping judgment: $\textsc{$\sub$-path}$. If
$p$ (path-)reduces to $q$, and $T \sub q.L$, then $T \sub p.L$. Path
reduction is a simplified form of reduction involving only
paths. However, this means that the subtyping judgment, and
indirectly, all the typing-related judgments, now need to carry the
store in addition to the context so that path reductions can be
calculated.

Now, let's see how path equality provisions and explicit widening can
fit together.

First, path reduction is not isomorphic to reduction anymore, since we
want to actually skip over widenings, as motivated by the last example
in~\ref{widening}.

In addition, we now also need a dual rule, $\textsc{path-$\sub$}$: if
$p$ (path-dually)-reduces to $q$, and $q.L \sub T$ then $p.L \sub
T$. This is because when we have a widening on an object on which a
method is called, we have to upcast the argument to the parameter type
expected by the original method. Here is a motivating example.

Let $T_c$ be a shorthand for the type:
\begin{align*}
\Top &\mlrefine z {\\
&\ \Ldecl A {\Top \refine z {\mdecl m \Bot \Top}} \Top\\
&\ \Ldecl B \Top \Top\\
&\ \mdecl m {z.A} \Top\\
}&
\end{align*}

Let $T$ be a shorthand for the type:
\begin{align*}
\Top &\mlrefine z {\\
&\ \Ldecl A {\Top \refine z {\mdecl m \Bot \Top}} \Top\\
&\ \Ldecl B \Top \Top\\
&\ \mdecl m {{z.A} \refine z {\Ldecl B \Top \Top}} \Top\\
}&
\end{align*}

Now, consider the term:
\begin{align*}
&\mlnew v {T_c \ldefs{m(x)=\wid x \Top}}{
(\wid v T).m(\wid v {((\wid v T).{A \refine z {\Ldecl B \Top \Top}})})
}
\end{align*}

When we evaluate the method invocation, we need to cast $\wid v
{((\wid v T).{A \refine z {\Ldecl B \Top \Top}})}$ to $v.A$, and for
this, we need the newly introduced $\textsc{path-$\sub$}$ rule.

Note that the path dual reduction can be a bit stricter with casts
than the path reduction. In any case, introducing this
$\textsc{path-$\sub$}$ rule into the subtyping judgment is
problematic: it is now possible to say $p.L \sub T$, even though $T$
can do more than what $p.L$ is defined to do. Here is an example,
where we construct an object, with $T = \Bot$. (The convolution in the
example is due to the requirement that concrete types be only
mentioned once.)

\begin{align*}
&\mlnew a {\Top \refine z {\Ldecl C \Bot {\Top \refine z {\Ldecl D \Bot {z.X}, \Ldecl X \Bot \Top}}}}{
\mlnew b {{a.C} \refine z {\Ldecl X \Bot \Bot}}}{
\mlnew c {a.C}}{
\mlnew d {(\wid b {a.C}).D}}{
\app {(\abs x \Bot {x.\mathit{foo}})} d
}
\end{align*}

Notice that $d$ has type $\Bot$ if you ignore the cast on $b$. This
example doesn't typecheck initially because $\textsc{path-$\sub$}$
only applies when objects are in the store, so the application is not
well-typed. But if we start preservation in a store which has $a$,
$b$, $c$ and $d$ then the application type-checks, because, through
$\textsc{path-$\sub$}$, we can find that the type of $d$ is a subtype
of $\Bot$. Now, of course, when we get to $d.\mathit{foo}$, reduction
fails.

So the preservation theorem as defined on a small-step semantics
(where we start with an arbitrary well-formed environment) fails when
we add the $\textsc{path-$\sub$}$ rule.

@@

Proving the DOT calculus type-safe has been an interesting
adventure. We have shown that DOT does not satisfy preservation (also
known as subject-reduction), because a well-typed term can step to an
intermediate term that is not well-typed. We're open to ideas on how
to resolve this. In any case, the standard theorems of preservation
and progress are just one way to prove type safety, which simply
states that a well-typed term cannot be stuck. We are developing a
proof of type safety using logical relations.
